---
# UW thesis fields
title: "Diabetes and Public Housing"
author: "Marc Macarulay"
year: "2020"
program: "Health Services"
chair:
chairtitle:
Committee :

abstract: |
  Public housing residents have worse health than the general population including chronic diseases such as diabetes. Knowledge of which specific populations within public housing residents have chronic conditions is essential for public health agencies to implement efficient health promotion programs. This cross-sectional design study was conducted using data collected from Data Across Sectors for Housing and Health partnership in King County, WA. Total population included were Medicaid and Medicare beneficiaries compiled from Washington Health Care Authority (n= 585,372). Associations between public housing and diabetes status were estimated using odds ratios and corresponding 95% confidence intervals from crude and adjusted models. Further associations were determined between individual public housing authorities and diabetes status using the same crude and adjusted models. Among study participants 10.4% received some form public housing assistance in 2017. In total, 9.9% were considered to have diabetes. The adjusted model revealed that public housing residents were 1.94 times (95% CI: 1.88-1.99) more likely to meet the diabetes definition compared to those not receiving housing assistance. These findings suggest that public housing residents are more likely to be diabetic. Further studies should explore the relationship between public housing and diabetes over a longer period to discern the direction of the association over time. Findings from this study can be used to inform future interventions for diabetes treatment by both public health and housing agencies.
acknowledgments: |
  "My acknowledgments"
# End of UW thesis fields
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output:
  huskydown::thesis_pdf:
    latex_engine: xelatex
#  huskydown::thesis_gitbook: default
#  huskydown::thesis_word: default
#  huskydown::thesis_epub: default
bibliography: bib/M.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: csl/apa.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
# space_between_paragraphs: true
# Delete the # at the beginning of the previous line if you'd like
# to have a blank new line between each paragraph
#header-includes:
#- \usepackage{tikz}
---

<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of metadata used to produce the document.  Be careful with spacing in this header!

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.
-->

<!--
If you receive a duplicate label error after knitting, make sure to delete the index.Rmd file and then knit again.
-->

```{r include_packages, include = FALSE}
# This chunk ensures that the huskydown package is
# installed and loaded. This huskydown package includes
# the template files for the thesis.
if(!require(devtools))
  install.packages("devtools", repos = "http://cran.rstudio.com")
if(!require(huskydown))
  devtools::install_github("benmarwick/huskydown")
library(huskydown)
```

```{r load-packages, include=FALSE}
library(dplyr)
library(magrittr)
library(knitr)
library(kableExtra)
library(xtable)
```
<!-- The {.unnumbered} option here means that the introduction will be "Chapter 0." You can also use {-} for no numbers on chapters. -->

<!--chapter:end:ThesisOutput.Rmd-->

<!--
The {#rmd-basics} text after the chapter declaration will allow us to link throughout the document back to the beginning of Chapter 1.  These labels will automatically be generated (if not specified) by changing the spaces to hyphens and capital letters to lowercase.  Look for the reference to this label at the beginning of Chapter 2.
-->
# Background and Significance {#rmd-basics}
## Public Housing
Housing is widely acknowledged as an important social determinant of health [@Thomson2013]. Health outcomes driven by housing are mediated by housing quality, safety, stability, and affordability [@Taylor2018]. There are well established links between housing quality and morbidity ranging from mental disorders, injuries, infectious diseases, and chronic diseases [@Krieger2002]. There is a growing body of evidence associating substandard and unstable housing with poor health outcomes [@Berkowitz2018 ; @Thomson2013]. However, less understood is the relationship between public housing and health.

The government tries to ameliorate housing problems by providing public housing assistance, which are decent and safe subsidized rental housing for eligible populations including low-income families, the elderly, and persons with disabilities [@HUD2020]. Public housing Authority (PHA) provides several types of housing subsidies that are low income public housing or voucher-based. A relevant study suggests that people with voucher-based subsidies that move from areas with a high level of poverty to areas with less poverty are associated with a lower prevalance of obesity and diabetes[@Ludwig2011]. Yet, other studies have shown that public housing residents have worse health outcomes than other city residents [@Digenis-Bury2008 ; @Manjarrez2007]. This relationship between public housing assistance in general and chronic conditions such as diabetes is minimally explored.

## Diabetes
Diabetes is a chronic disease that is characterized by an inability of the body to maintain a healthy blood glucose level, this can cause a variety of symptoms that affect multiple systems in the body and can lead to potentially life-threatening complications. The key regulator hormone of glucose is insulin and it is produced in the pancreas. The absence or malfunction of insulin leads to elevated blood glucose levels called hyperglycemia. When insulin hormone is missing or ineffective the disease is called Diabetes Mellitus and this condition has multiple types.

### Diabetes Variants
The most common diabetes variants include type I diabetes mellitus, type II diabetes mellitus, and gestational diabetes. Type I diabetes is usually caused by genetic factors triggering an autoimmune reaction that results in the destruction of insulin producing cells in the pancreas. Also known as Juvenile Diabetes, the type I classification is typically diagnosed relatively early in life during childhood or early adulthood. Whereas Type II diabetes develops when the body can still produce insulin however the amount is insufficient or when the body becomes resistant to the effects of insulin. Type II diabetes is largely attributed to lifestyle factors. Gestational diabetes is the least common type and occurs during pregnancy. The prevalence of type II diabetes is much higher than type I. In the US, type II and type I diabetes account for approximately 91% and 6% of all diagnosed diabetes cases, respectively [@Bullard2018].

Diabetes is a serious chronic condition without a medical cure. The treatment for diabetes involves disease prevention and management. Medical treatment of diabetes primarily consists of exogenous insulin replacement or use of medications that stimulate the pancreas to produce endogenous insulin. Without adequate blood control, diabetes can lead to increased risk of other conditions including vision loss, heart disease, stroke, kidney failure, nerve damage, amputation, and even premature death.

##Problem Definition
Disease management for type II diabetics focuses on lifestyle modification such as diet control and increased physical activity. The goal is to promote weight loss and reduce excess fat that subsequently both reduces insulin resistance and enhances disease control. However, other determinants of health have been recognized to impact the effectiveness of diabetes management, namely healthcare access, cultural and social support, economic stability and built environments [@Clark2014]. Housing instability and food insecurity in particular have been shown to reduce diabetes management self-efficacy for low income adults [@Vijayaraghavan2011]. Providing safe and affordable housing to those in need may improve overall health and increase diabetes management efficacy.

While there is a growing number of published literature on the correlation between substandard housing and health outcomes, few studies specifically examine the relationship between general public housing support and diabetes status. For this reason, the current study aimed to explore this public health issue within a local context in King County, WA.

In the effort to decrease the gap of knowledge between the junction of public housing and health, Public Health Seattle and King County (PHSKC) developed a unique partnership with King County Housing Authority (KCHA) and Seattle Housing Authority (SHA). This partnership enabled data to be shared across sectors with the intention of informing and measuring future interventions that would improve the health of the county residents. This project expands on the original research from this collaboration revolving around health and public housing. This current research aims to use the provided data to contribute to the literature on the association between public housing and diabetes among Medicaid and Medicare patients.

The findings of the study could help identify where resources for diabetes prevention and management might be used more effectively. For instance, PHAs can look into directly implementing diabetes prevention programs within PHA run housing communities to reduce the burden of diabetes. Furthermore, identifying populations with higher needs of diabetes related care can be used as a metric for implementing health policy.    

<!--chapter:end:Background.Rmd-->

# Methods {}
## Study Setting and Study Design
The current study investigates whether public housing is associated with risk of diabetes status among King County, WA residents who were enrolled in Medicare and Medicaid. This study uses a descriptive cross-sectional design. The cross-sectional design is appropriate because it allows for an estimate of a dichotomous disease outcome at a particular point in time [@Greenland1988].

The analysis of this study was conducted on a dataset compiled from the King County _Data Across Sectors for Housing and Health (DASHH)_ partnership. The findings from the original initial study have previously been reported [@PHSKC2018].

## Data Sources
In an effort to reduce fragmented data siloes across different sectors, the DASHH partnership was formed in 2016 between Public Health - Seattle and King County, and two public housing authorities, King County Housing Authority and Seattle Housing Authority. The primary objectives for DASHH were to join health and housing administrative data together to inform and measure future interventions, relating to policy, outreach, and program evaluation that would improve the health of King County residents, as well as to disseminate actionable data to key health and housing stakeholders.

KCHA provides housing assistance to families and individuals that earn less than the county median income to approximately 55,000 people in the King County area excluding the cities of Seattle and Renton. While SHA predominantly focuses on people within Seattle, with nearly 35,000 people scattered throughout 400 sites across the city.

The housing data originated from both KCHA and SHA extracted from a standardized US Department of Housing and Urban Development (HUD) form. This data source contained elements that included demographic information and period of enrollment for families and individuals. Claims and enrollment data for Medicaid and Medicare were provided by the Washington Health Care Authority (HCA) and the WA Office of Financial Management, respectively. Enrollment data contained information on who was receiving health insurance benefits. Claims data provided elements such as diagnosis codes and these were used to identify acute events and chronic conditions. The separated health and housing data sources were linked together probabilistically to create a unique identifier. Further detail on the process of joining data can be found in the King County Data Across Sectors for Housing and Health, 2018 Report [@PHSKC2018]. This report provides a detailed technical process of linking Medicaid data in the orignal study, but this approach was similar to both the Medicaid and Medicare data used for this study.

## Study Population
The study population was participants that were enrolled in either Medicare or Medicaid programs. Further eligibility for study participation included King County, Washington residency and at least 11 months of Medicare or Medicaid coverage in 2017. The 11 month minimum coverage restriction provides a more complete picture of healthcare usage in 2017, reducing the artificial inflation of the eligible population with people that had little chance of meeting the diabetes definition. The overall number of participants derived from the DASHH dataset totaled 585,372.

### Exposure Variable
The exposure variable for this study was public housing assistance status. This was extracted from the HUD-50058 form which was provided by the PHAs. The HUD-50058 form provides information on families that participate in public housing or Section 8 rental subsidy programs [@HUD2020]. Housing assistance is separated into 3 main types:

* Housing Choice Vouchers - vouchers provided to recipients to rent units on the private housing market
* Public housing properties and units - subsidized housing managed by PHAs
* Project-based vouchers - subsidized housing units not managed by PHAs

Responses on the HUD-50058 form were combined into a composite public housing binary variable. Study participants that were not enrolled in any of the listed housing assistance programs were not placed in PHA, whereas any of the 3 types of housing assistance were categorized into the PHA group.

### Outcome Variable
The outcome variable for this study was diabetes status. This was defined using the Centers for Medicare and Medicaid Services (CMS) Chronic Conditions Warehouse (CCW) algorithm [@CMS2020]. According to the CCW, a participant meets the criteria if they have at least 1 inpatient, skilled nursing facility, home health agency visit or 2 hospital outpatient or carrier claims with diabetes diagnoses codes as outlined by the chronic conditions reference list within the last 2 years[@CMS2020]. This definition does not specify diabetes variant but instead accounts for any type diabetes diagnoses. The diabetes status outcome variable was dichotomous.

### Potential Confounders
Potential confounders were identified based on literature review. This study considers age, race and ethnicity and gender as potential confounding variables. Each of these variables were selected due to the increased baseline risk for participants to be either in public housing or have diabetes. It is known that diabetes is an age-related disease, with a higher risk for older populations [@Selvin2013]. Age was presented as a discrete variable for the participants age in 2017. Similarly, according to CDC data, racial minority groups may be differentially at risk for both type 1 and type 2 diabetes compared to their white counterparts [@Divers2020 ; @CDC2020]. Although the discrete variable for age was used in the statistical analyses, age is presented categorically in the descriptive analyses for a simpler distribution description. Mean and median age were also shown for each category. Race and ethnicity variable was defined categorically and included: American Indians/Alaska Natives, Asian, Asian Pacific Islander, Black/African American, Latino, Multiple, Native Hawaiian and Pacific Islander, Other, Unknown, and White. Due to the limitations of the how Medicare data categorizes race, this study categorized Asian Pacific Islander as a distinct group from the other separate groups of Asian and Native Hawaiian and Pacific Islander. Gender was selected because both psychosocial and biological factors are responsible for sex and gender diabetes risk differences [@Kautzky-Willer2016]. Gender was grouped categorically and included: Female, Male, and Multiple.

## Analyses
For the statistical analyses, logistic regression models were fitted to assess the risk of diabetes status in relation to public housing assistance status. This analysis is well-suited for this study because logistic regression analyses allows for measuring the association of an effect towards a binomial response variable by combining multiple variables to avoid confounding [@Sperandei2014].

There were two main research questions in this study. The primary research question delved into the relationship between public housing assistance and diabetes. The secondary question looked into the relationship between each of the public housing authorities and diabetes status. The distinction between these analyses are that first provides the likelihood of meeting the diabetes definition if a person was receiving public housing assistance and the second provides a more specific estimate for each of the housing authorities.

Crude and adjusted models were fitted to answer the research questions. The adjusted model accounted for age, race and ethnicity and gender variables. The models were used to determine the odds ratios (OR) and corresponding 95% confidence intervals. Findings were statistically significant if the estimates did not cross the confidence intervals and p-values were below <0.05 threshold. Analyses were conducted using statistical software R version 3.6.0 and RStudio version 1.2.

<!--chapter:end:Methods.Rmd-->


# Results {#ref-labels}
## Descriptive Statistics
Among King County residents on Medicaid or Medicare in 2017, 10.4% also received support from a PHA (5.9% were in KCHA and 4.6% in SHA) (See Table \@ref(tab:table1)). A greater proportion of people in the PHA population met the definition of diabetes at 12.7% compared to the non-PHA group with 9.6%. Overall, 9.9% met the definition of diabetes. Additionally, the population age distributions were different between PHA status; the non-PHA category had an older population (median of age of 62.0, mean age of 50.0 compared with the PHA population median and mean age of 34.0 and 35.7, respectively). The PHA group were more racially and ethnically diverse however, the gender distribution between the two groups were similar.

## Public Housing and Diabetes
In the primary analysis, the association between diabetes status and public housing assistance, the crude model showed that the odds ratio of having diabetes was 1.34 fold greater for those receiving public housing assistance (See Table \@ref(tab:table2)). This effect increased in the adjusted model when the potential confounders were added, PHA residents were 94% more likely to meet the definition of diabetes compared to those that were non-PHA residents (See Table \@ref(tab:table3)).

Housing Status | Unadjusted        | Adjusted                
---------------|------------------|-----------------               
Non-PHA        | Reference         | Reference                    
PHA            | 1.34 (CI: 1.31-1.38) | 1.94 (CI: 1.88-1.99)    
Table: (\#tab:PHA) Association between PHA Status and Diabetes

Housing Status  | Unadjusted             | Adjusted
--------|----------------------|---------------------
Non-PHA | Reference             | Reference                                                            
KCHA    | 1.28 (CI: 1.24-1.33) | 2.16 (CI: 2.09-2.25)
SHA     | 1.42 (CI: 1.38-1.48) | 1.70 (CI: 1.64-1.77)
Table: (\#tab:PHAA) Association between the Public Housing Authorities and Diabetes  

## Public Housing Authorities and Diabetes
In the secondary analysis, estimating the association between diabetes status and the specific public housing authorities, the crude model showed that the odds of meeting the definition of diabetes were 1.28 times greater among KCHA residents and 1.42 times greater among SHA residents (See Table \@ref(tab:table4)). The adjusted model revealed that among KCHA residents, the odds of meeting the definition of diabetes were 2.16 times higher and 1.70 for SHA residents compared to non-PHA residents (See Table \@ref(tab:table5)).

<!--chapter:end:Results.Rmd-->

# Discussion {}
##Discussion
Findings from this study indicate that, after adjusting for age, gender and race and ethnicity, public housing assistance was positively associated with diabetes status. The findings also suggest that the increased burden of diabetes is not evenly spread between the two PHAs included in the study.

The increased burden of diabetes among those supported by PHAs observed in this study reinforces the results to a similar study that compared public housing residents in Boston, MA to other city residents [@Digenis-Bury2008]. A likely reason for these findings is that PHAs specifically target individuals with high-needs when allocating housing resources, this results in an overall sicker PHA population from the beginning. The difference in the burden of diabetes between KCHA and SHA may be attributed to the proportion of people with high-needs that are given priority for scarce housing resources. An additional factor that may contribute to these findings is that public housing residents were more likely to be racial and ethnic minorities and the prevalence of diabetes is often greater for racial and ethnic populations [@Chow2012]. Another possible explanation is that areas and neighborhoods with a higher level of poverty, such as where public housing properties may be located tend to also have a higher prevalence of obesity and diabetes [@Ludwig2011].

These findings suggest that public housing assistance recipients have a greater need for diabetes treatment and programs run by PHAs could be an avenue in which resources on diabetes prevention and management may be more effective. A health promotion intervention that utilized homecare nurses to implement a diabetes prevention program in public housing communities proved to be successful and can be a viable option in improving health outcomes [@Whittemore2014]. Considering the population demographics in the PHA group, another successful strategy may be to implement a culturally competent diabetes care intervention program [@Zeh2012].

However, the preliminary findings in this study indicating that PHA recipients are more likely to be diabetic are inferential and therefore cannot be interpreted as a causal relationship. As previously mentioned, substandard and unstable housing has been linked with worse diabetes health outcomes, however stable housing provided by PHA has been associated with the ability for participants to afford diabetes-related financial expenses [@Keene2018]. Additionally, when housing needs are met participants are able to prioritize diabetes self-management, avoiding potential complications [@Keene2018]. Participants with diabetes prior to receiving public housing assistance could potentially see a reduction in diabetes related complications over time. Furthermore, housing mobility, a strategy that helps low income people move to neighborhoods with better resources through housing vouchers has proven to be a successful policy in terms of improved health outcomes [@Gale2018]. Advocating for policies that increase housing mobility among high-needs populations could help to address the burden of diabetes and other health outcomes.

Strengths of this study include the population-based study design that allows for estimation of disease impact on a broad scale and the well-defined study population. The data used in this study were also from reliable sources with careful and normally complete documentation, subsequently reducing the impact of information bias.

## Limitations
There are several limitations to note. First, there may have been unmeasured potential confounders that have not been included in the analyses. Other elements that are recognized to be potential confounders for diabetes may have been useful to include in this study like socioeconomic status and other health characteristics like BMI. Given the obtained data, these elements were not available and thus not accounted for in this study.      

While this study examines public housing assistance in general, this study does not specifically look into the public housing living standards. The living conditions of the public housing support may vary from unit to unit especially among voucher-based recipients. Adjusting for this variation would provide more precise estimates considering how disease management is affected by housing conditions.

Another notable limitation is that due to the inherent characteristic of claims data, the population captured in the study were only those that sought healthcare services for diabetes care or diabetes related outcomes. People who may have been diabetic during this period but were asymptomatic or those who had been previously diagnosed with diabetes but did not seek care within the time frame provided by the CCW algorithm definition of diabetes were not captured in the study. Consequently, the eligible population in this study cannot provide a prevalence estimate by themselves.

Despite the limitations, this study contributes to our understanding of government subsidized housing and diabetes self-management. The findings are generalizable to low-income, racially, and ethnically diverse populations with diabetes who obtain health care in safety-net health settings. More specifically, these findings provide estimates for those with diabetes in public housing and can be used to inform and evaluate existing interventions as well as create new strategies for health promotion within this population. Future studies should continue investigating the relationship between public housing and diabetes and in particular further studies should explore the association over a longer time period.

<!--chapter:end:Discussion.Rmd-->

`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'`

<!--
If you feel it necessary to include an appendix, it goes here.
-->

#Appendix {-}

```{r table1, echo=FALSE}
options(knitr.kable.NA = '')
table1<-readr::read_csv("table 1.csv", col_types = readr::cols())
table1%>%
knitr::kable(format = "latex", booktabs= T, caption= "Population Demographics")%>%
kable_styling(font_size =  12 , full_width = TRUE)%>%
  pack_rows("Age",2,11, bold= TRUE )%>%
  pack_rows("Race and Ethnicity",12,21, bold= TRUE)%>%
  pack_rows("Gender",22,24, bold= TRUE)%>%
  column_spec(1,width = "6.1cm")%>%
  footnote(general = "Percentages may not add up to 100 because of missing data")

```


<br />
<br />
```{r table2, echo=FALSE}
table2<-readr::read_csv("diabetes_crude_test_Output.csv", col_types = readr::cols())
table2%>%
knitr::kable(format = "latex", booktabs= T, caption= "Crude PHA Regression Model")%>%
  kable_styling(font_size =  12 , full_width = T)%>%
  footnote(general = "Reference groups were Female and AI/AN"

```

```{r table3, echo=FALSE}
table3<-readr::read_csv("Adjusted_Diabetes_PHA_Output.csv", col_types = readr::cols())
table3%>%
knitr::kable(format = "latex", booktabs= T, caption= "Adjusted PHA Regression Model")%>%
  kable_styling(font_size =  12 , full_width = T)
```

```{r table4, echo=FALSE}
table4<-readr::read_csv("pha_agency_output.csv", col_types = readr::cols())
table4%>%
knitr::kable(format = "latex", booktabs= T, caption= "Crude PHA Agency Regression Model")%>%
  kable_styling(font_size =  12 , full_width = T)
```


```{r table5, echo=FALSE}
table5<-readr::read_csv("pha_agency_adj_output.csv", col_types = readr::cols())
table5%>%
knitr::kable(format = "latex", booktabs= T, caption= "Adjusted PHA Agency Regression Model")%>%
  kable_styling(font_size =  12 , full_width = T)%>%
  footnote(general = "Reference groups were Female and AI/AN"
```

<!--chapter:end:Appendixes.Rmd-->

<!--
The bib chunk below must go last in this document according to how R Markdown renders.  More info is at http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
-->

\backmatter

<!-- 
If you'd like to change the name of the bibliography to something else,
delete "References" and replace it.
-->

# References {-}
<!--
This manually sets the header for this unnumbered chapter.
-->
\markboth{References}{References}
<!--
To remove the indentation of the first entry.
-->
\noindent

<!--
To create a hanging indent and spacing between entries.  These three lines may need to be removed for styles that don't require the hanging indent.
-->

\setlength{\parindent}{-0.20in}
\setlength{\leftskip}{0.20in}
\setlength{\parskip}{8pt}


<!--
This is just for testing with more citations for the bibliography at the end.  Add other entries into the list here if you'd like them to appear in the bibliography even if they weren't explicitly cited in the document.
-->

---
nocite: | 
  @angel2000, @angel2001, @angel2002a
...



<!--chapter:end:99-references.Rmd-->

